---
title: "Project 4 - Oregon Election Contributions 2016"
author: "Sherry Parker"
date: "March 30, 2016"
output: html_document
---



# Project 4 - Evaluate Oregon Contributions 
&nbsp;
#           to the 2016 Presidential Election
&nbsp;
#            April 1, 2015 through Sept 29, 2015
***

## Data Information
#Notes: The data runs from April 1, 2015 through Sept 29, 2015 and includes all
#candidates
 
  
```{r libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
library(RColorBrewer)
```

```{r create palletes}
#color blind friendly pallete from 
#http://www.cookbook-r.com/Graphs/Colors_%28ggplot2%29/
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                "#0072B2", "#D55E00", "#CC79A7")

#Color palette for Political Party
party_color <- c("DEMOCRAT" = "blue", "REPUBLICAN" =  "red")

#Color palette for Political Party
gen_color <- c("MALE" = "chartreuse4", "FEMALE" =  "darkorchid1")
```


```{r import data}
#Import Contributions for 2016 Election from Oregon
cont_table <- read.csv("P00000001-OR.csv")

#Basics data info
#Added Candidate Political Party, Candidate Gender, and County columns to the
#original data set.

#Adjust date information to Year-Month combo for grouping purposes
cont_table$contb_receipt_dt <- as.Date(cont_table$contb_receipt_dt, format="%d-%b-%y")
cont_table$mth_yr <- format.Date(cont_table$contb_receipt_dt, format="%b-%y")
cont_table$mth_yr <- factor(cont_table$mth_yr)

dim(cont_table)
summary(cont_table)
sapply(cont_table, class)

first_date <- min(cont_table$mth_yr)
last_date <- max(cont_table$mth_yr)

first_date
last_date

#Lets wipe out the ones that are before 2015 and see what is left

cont_table <- subset(cont_table, format(contb_receipt_dt, '%Y')=='2015')

#verify date changes to master table
first_date <- min(cont_table$contb_receipt_dt)
last_date <- max(cont_table$contb_receipt_dt)

first_date
last_date                        

#Yes that did the job correctly
```

***

## Evaluate data by Candidate
Looking to see any patterns, trends, or anomolies that arise

```{r Evaluate by Candidate}
candidate_groups = group_by(cont_table, cand_nm)
total_contb_amt <- sum(cont_table$contb_receipt_amt)
total_n_contb <- max(cont_table$index)
total_n_contb

#Summary of contribution value statistics by candidate
cont_table.fc_by_cand = summarise(candidate_groups,
                             cand_mean = mean(contb_receipt_amt),
                             cand_median = median(contb_receipt_amt),
                             cand_sum = sum(contb_receipt_amt),
                             cand_sum_pct = (cand_sum/total_n_contb),
                             n = n(),
                             cand_n_pct = (n/total_n_contb))

cont_table.fc_by_cand <- arrange(cont_table.fc_by_cand, desc(cand_mean))

cont_table.fc_by_cand2 <- arrange(cont_table.fc_by_cand, desc(n))

###Boxplot of contributions per candidate
#help from http://stackoverflow.com/questions/1330989/rotating-and-spacing-axis
#-labels-in-ggplot2
p1 <- ggplot(aes(x = cand_nm, y = contb_receipt_amt), data = cont_table) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Contributions by Candidate - BoxPlot")
p1
#I find it interesting that some of the box plots show a greater spread than 
#others do. Also this shows that there are negative values which are unexpected.
```
***

```{r Evaluate by Candidate2}
###Do candidates that have a smaller mean contribution have more contributors
#(mean contribution vs # of contributions)
#Chart of mean contributions compared to number of contributors

p2 <- ggplot(aes(x = cand_mean, y = n, color = cand_nm), 
             data = cont_table.fc_by_cand) +
  geom_point() +
  ggtitle("Mean Contributions vs Number of Contributions")
p2

# There appears to be a relationship between contribution size and number 
#of contributors but it is not linear. Also with so many candidates it is hard 
```
***

```{r Eval by Cand3}

#to know which color goes with who.  Looking at those under $200 mean should 
#help that some.


p2a <- ggplot(aes(x = cand_mean, y = n, color = cand_nm), 
              data = subset(cont_table.fc_by_cand, 
                            (cand_mean <= 200 & cand_mean > 0))) +
  geom_point() +
  scale_color_manual(values = cbbPalette,
                     guide = guide_legend(title = 'Ccandidated', 
                                          reverse = T,
                                          override.aes = list(alpha = 1, 
                                                              size = 2))) +
  ggtitle("Mean Contributions vs Number of Contributions")
p2a
#It would appear that there is no direct correlation
```
***


## Total contributions by candidate

```{r Eval by Cand 4}
#Plotted as a bar chart
p3 <- ggplot(aes(x = cand_nm, y = cand_sum), data = cont_table.fc_by_cand) +
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Total Contributions by Candidate")
p3

#Stacked Bar Chart
p3a <- ggplot(aes(x = "", y = cand_sum_pct, fill = cand_nm), 
              data = cont_table.fc_by_cand2) +
  geom_bar(width = 1, stat = "identity")
p3a

#Pie Chart
p3b <- p3a + coord_polar("y", start=0)
p3b
#There is something off on the percentages as it goes past 200%
```
***

## Top 5 Candidates by Total Contributions

```{r top 5 cand}
#top 5 by contributions
head(cont_table.fc_by_cand2, 5)

#Set of the top 5 candidates to use for later analysis
top_five = subset(cont_table, cand_nm == "Sanders, Bernard" |
                    cand_nm == "Carson, Benjamin S." |
                    cand_nm == "Clinton, Hillary Rodham" |
                    cand_nm == "Cruz, Rafael Edward 'Ted'" |
                    cand_nm == "Paul, Rand")
```

## Evaluate by Political Party
Looking to see any patterns, trends, or anomolies that arise

```{r Evaluate by Political Party}


#Do Oregonions support Democrats or Republicans more 
#(Total contributions by party)

party_group = group_by(cont_table, cand_party)

cont_table.fc_by_party = summarise(party_group,
                                  party_mean = mean(contb_receipt_amt),
                                  party_median = median(contb_receipt_amt),
                                  party_sum = sum(contb_receipt_amt),
                                  n = n())

cont_table.fc_by_gen



p3c <- ggplot(aes(x = cand_party, y = party_sum,
                  fill=cand_party), 
              data = cont_table.fc_by_party) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values=party_color) +
  ggtitle("Total Contributions by Party")
p3c
#I am surprised to see the total contributions between the two parties to be
#so close.  I guess living in the Urban "blue" part of the state jaded my
#expectations.


```
## Evaluate by Candidate Gender
Looking to see any patterns, trends, or anomolies that arise

```{r Evaluate by Candidate Gender}


#Do Oregonions support Female or Male Candidates more 
#(Total contributions by gender of candidate)

gen_group = group_by(cont_table, cand_gen)

cont_table.fc_by_gen = summarise(gen_group,
                                  gen_mean = mean(contb_receipt_amt),
                                  gen_median = median(contb_receipt_amt),
                                  gen_sum = sum(contb_receipt_amt),
                                  n = n())





p3d <- ggplot(aes(x = cand_gen, y = gen_sum,
                  fill=cand_gen), 
              data = cont_table.fc_by_gen) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values=gen_color) +
  ggtitle("Total Contributions Grouped by Candidate Gender")
p3d

```
***

## Evaluate zip code
Looking to see any patterns, trends, or anomolies that arise

```{r Evaluate by zip code}

#What about by zip code (total contributions by party colored party 
#and faceted by candidate)
#help from http://stackoverflow.com/questions/33199203/
#r-how-to-display-the-first-n-characters-from-a-string-of-words

#create column for 5 digit zip to consolidate
cont_table$zip5 = substr(cont_table$contbr_zip, start=1, stop=5)

zip_group = group_by(cont_table,zip5)

cont_table.fc_by_zip = summarise(zip_group,
                                  zip_mean = mean(contb_receipt_amt),
                                  zip_median = median(contb_receipt_amt),
                                  zip_sum = sum(contb_receipt_amt),
                                  n = n())

p4 <- ggplot(aes(x = zip5, y = zip_sum), 
             data = cont_table.fc_by_zip) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Total Contributions by 5-digit Zip Code")
p4
#Too many for this to be very meaningful.  Let's see what the top 10
#are and graph just those.
```

```{r top 5 zip}

cont_table.fc_by_zip <- arrange(cont_table.fc_by_zip, desc(zip_sum))

head(cont_table.fc_by_zip, 5)

cont_table.fc_by_zip_top5 <- subset(cont_table,
                                    zip5 == "97201" |
                                    zip5 == "97210" |
                                    zip5 == "97219" |
                                    zip5 == "97221" |
                                    zip5 == "97405")

p4a <- ggplot(aes(x = zip5, y = zip_sum), 
             data = cont_table.fc_by_zip_top5) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Total Contributions by 5-digit Zip Code")
p4a
```

```{r top 5 zip by party}

#What about if we color by party?

p4b <- ggplot(aes(x = cand_party, y = contb_receipt_amt, 
                  fill = cand_party), data = cont_table.fc_by_zip_top5) +
  geom_bar(stat = "identity") +
  facet_wrap( ~ zip5) +
  scale_fill_manual(values=party_color) +
  ggtitle("Top 5 Zip Codes total contributions by party")
p4b

#Zip code 97201 is much of what is considered downtown Portland

```
***
## Evaluate Candidate and Party through time
Looking to see any patterns, trends, or anomolies that arise

```{r Evaluate by Candidate and Party over time}
#Create a data frame to sum by month for time related graphing
cont_table.fc_by_summarize  <- 
  cont_table  %>% 
  group_by(cand_nm, mth_yr, cand_gen, cand_party) %>%
  summarise(monthly_total = sum(contb_receipt_amt))

mth_sort = factor(cont_table.fc_by_summarize$mth_yr, levels = c("Apr-15","May-15","Jun-15","Jul-15","Aug-15","Sep-15"))
mth_sort



p5 <- ggplot(aes(x = mth_sort, y = monthly_total, 
                 color= cand_party), 
             data = cont_table.fc_by_summarize)+
  geom_point() +
  facet_wrap( ~ cand_nm) +
  scale_color_manual(values=party_color) +
  scale_y_continuous(limits = c(0, 3000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Contributions through time - All Candidates")
p5

#Too many dates to be meaningful, plus there are some odd dates back in 2013.
#Lets reset the main table to only show 2015 and sooner


```

## Evaluate Democrats only

```{r Evaluate the Democrats over time}

#comparison of just the democratic contributions over time with help from
#http://docs.ggplot2.org/current/scale_date.html
p6 <- ggplot(aes(x = contb_receipt_dt, y = contb_receipt_amt, 
                 color = cand_nm), 
                   data = subset(cont_table, 
                                 contb_receipt_amt > 0 & 
                                   cand_party == "DEMOCRAT")) +
  geom_point() +
  facet_wrap(~ cand_nm) +
  scale_y_continuous(limits = c(0, 3000),
                     breaks = c(0, 500, 1000, 1500, 2000, 2500, 3000)) +
  ggtitle("Democrate contributions through time - Democrats")
p6

#This chart shows that people prefer to contribute at round numbers like $500, 
#or $1000.  you can also see that there are a number at $2700 which is the max 
#contribution. Lots of noise below $250 so digging deeper.
```

```{r dems over time2}
#What about in the busy lower range under $300
p6a <- ggplot(aes(x = contb_receipt_dt, y = contb_receipt_amt, 
                  color = cand_nm), 
                   data = subset(cont_table, 
                                 contb_receipt_amt > 0 & 
                                   cand_party == "DEMOCRAT")) +
  geom_point() +
  facet_wrap(~ cand_nm) +
  scale_y_continuous(limits = c(0, 300),
                     breaks = c(0, 50, 100, 150, 200, 250, 300)) +
  ggtitle("Democrate contributions through time - Democrats")
p6a
#Still see distinct lines at round numbers, which isn't really all 
#that surprising.
```

```{r dems over time 3}
#What about the busy at the bottom still
p6b <- ggplot(aes(x = contb_receipt_dt, 
                  y = contb_receipt_amt, 
                  color = cand_nm), 
                   data = subset(cont_table, 
                                 contb_receipt_amt > 0 & 
                                   cand_party == "DEMOCRAT")) +
  geom_point() +
  facet_wrap(~ cand_nm) +
  scale_y_continuous(limits = c(0, 100),
                     breaks = c(0, 25, 50, 100)) +
  ggtitle("Democrate contributions through time - Democrats")
p6b
#Noticeably more donation for Sanders at this level


```
***
## Evaluate Republicans Only

```{r Evaluate the Republicans over time}
p7 <- ggplot(aes(x = contb_receipt_dt, y = contb_receipt_amt, 
                 color = cand_nm), 
             data = subset(cont_table, 
                           contb_receipt_amt > 0 & 
                             cand_party == "REPUBLICAN")) +
  geom_point() +
  facet_wrap(~ cand_nm) +
  scale_y_continuous(limits = c(0, 3000),
                     breaks = c(0, 500, 1000, 1500, 2000, 2500, 3000)) +
  ggtitle("Republican contributions through time")
p7

#You see the same pattern of round value contributions in the republicans too.
#Same visibilty issues on the low ranges, lets break this out more too
```

```{r Republicans over time 2}
p7a <- ggplot(aes(x = contb_receipt_dt, 
                  y = contb_receipt_amt, 
                  color = cand_nm), 
             data = subset(cont_table, 
                           contb_receipt_amt > 0 & 
                             cand_party == "REPUBLICAN")) +
  geom_point() +
  facet_wrap(~ cand_nm) +
  scale_y_continuous(limits = c(0, 300),
                     breaks = c(0, 50, 100, 150, 200, 250, 300)) +
  ggtitle("Republican contributions through time")
p7a
#Wow, lots of chatter in the data for Carson and Cruz.  
#Also several with nothing in this spot, should probably take them out.
```

```{r republicans over time 3}
#Break it down again
p7b <- ggplot(aes(x = contb_receipt_dt, 
                  y = contb_receipt_amt, 
                  color = cand_nm), 
             data = subset(cont_table, 
                           contb_receipt_amt > 0 & 
                             cand_party == "REPUBLICAN")) +
  geom_point() +
  facet_wrap(~ cand_nm) +
  scale_y_continuous(limits = c(0, 100),
                     breaks = c(0, 25, 50, 75, 100)) +
  ggtitle("Republican contributions through time")
p7b
#3 more with no data.  Surprised to see Trump on the lists here as I 
#understood he was not taking donations
```
***
## All candidates over time

```{r candidates over time}

#sum of donations over time colored by candidate
time_cand_group = group_by(cont_table, mth_yr, cand_nm)



all_by_month = aggregate(. ~ mth_yr,cand_nm, data=cont_table, FUN=sum)

cont_table.fc_by_date = summarise(time_cand_group,
                                  dt_cand_sum = sum(contb_receipt_amt))

all_by_month



p8 <- ggplot(aes(x = contb_receipt_dt, y = dt_cand_sum, color = cand_nm), 
           data = cont_table.fc_by_date) +
  geom_point() +
  ggtitle("Total Contributions through time")
p8
#Lots of noise at the bottom, time to look at less then $500 a day, also need 
#to fix the dates on the y axis
```

```{r cands over time 2}

p8a<- ggplot(aes(x = contb_receipt_dt, 
                 y = dt_cand_sum, 
                 color = cand_nm), data = cont_table.fc_by_date) +
  geom_point() +
  scale_y_continuous(limits = c(0, 500),
                     breaks = c(0, 100, 200, 300, 400, 500)) +
  ggtitle("Total Contributions through time")
p8a
#Can still see straight lines at round numbers, but there is far too much 
#chatter in this to be a worthwhile chart
```

```{r top 5 republicans over time}
#What if we only look at the top candidates
time_cand_group_five = group_by(top_five, contb_receipt_dt, cand_nm)

cont_table.fc_by_date_five = summarise(time_cand_group_five,
                                  dt_cand_sum = sum(contb_receipt_amt))

p8b<- ggplot(aes(x = contb_receipt_dt, y = dt_cand_sum, color = cand_nm), 
             data = cont_table.fc_by_date_five) +
  geom_point() +
  scale_y_continuous(limits = c(0, 500),
                     breaks = c(0, 100, 200, 300, 400, 500)) +
  ggtitle("Total Contributions through time - Top 5 Candidates by Donations")
p8b
```

## Political Party Factored by County

```{r by party factored by county}
#Total Contributions for each party factored by zip code
p9 <- ggplot(aes(x = cand_party,  y = contb_receipt_amt, fill = cand_party), 
             data = subset(cont_table, contb_receipt_amt >0)) +
  facet_wrap(~county) +
  scale_fill_manual(values=party_color) +
  geom_bar(stat = "identity", position = "identity")
p9



# This makes it clear that Multnomah, Clackamas and Washington Counties are 
#the counties in the race.  considering these are the most populous areas I am 
#not surprised. Also of interest is that some counties show no donations to 
#demoracts at all.  These are all rural counties. Just to be sure lets confirm 
#the top 3.

county_group = group_by(cont_table, county)

cont_table.fc_by_county = summarise(county_group,
                             county_mean = mean(contb_receipt_amt),
                             county_median = median(contb_receipt_amt),
                             county_sum = sum(contb_receipt_amt),
                             n = n())

cont_table.fc_by_county <- arrange(cont_table.fc_by_county, desc(county_sum))
```

```{r top cand by county }
#Lets graph it

p9aa <- ggplot(aes(x = county,  y = county_sum), 
               data = cont_table.fc_by_county) +
  geom_bar(stat = "identity", position = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Total Contributions by County")

p9aa
#Not at all surprising considering that the Portland Area accounts for a 
#significant portion of Oregon's population.

top_counties <- head(cont_table.fc_by_county, 3)
top_counties

#Lets look at just the heavy hitters, which are not surprisingly the 
#population dense urban and suburban counties around the Portland 
#Metrolopolitan area.

lead_counties = c(county == "Clackamas County" |
                 county == "Multnomah County" |
                 county == "Washington County")


cont_lead_counties = subset(cont_table, county == "Clackamas County" |
                 county == "Multnomah County" |
                 county == "Washington County")



lead_counties_five = subset(cont_lead_counties, cand_nm == "Sanders, Bernard" |
                    cand_nm == "Carson, Benjamin S." |
                    cand_nm == "Clinton, Hillary Rodham" |
                    cand_nm == "Cruz, Rafael Edward 'Ted'" |
                    cand_nm == "Paul, Rand")


```
# Final Evaluation Graphs
These charts are the final charts showing more detailed analysis

``` {r final evaluation graph 1}

#And now those three top counties by the top candidates.
#Lets look at these counties by party, I would expect to find a lean towards 
#democrat especially in Multnomah county
p9a <- ggplot(aes(x = cand_party,  y = contb_receipt_amt, fill = cand_party), 
              data = subset(cont_lead_counties, contb_receipt_amt > 0)) +
  facet_wrap( ~ county) +
    scale_fill_manual(values=party_color) +
  geom_bar(stat = "identity", position = "identity") +
  ggtitle("Three Counties with Highest Total Donation Dollars")
p9a

#As expected there is more money coming in from Multnomah County compared
#to any of the others.  It is suprising for an area so often considered
#Blue that there is still considerable Republican donations.
```

``` {r final eval graph 2}

#Graph showing the same three counties, divided by candidate and 
#indicating parties.

p9b <- ggplot(aes(x = cand_nm,  y = contb_receipt_amt, fill = cand_party), 
              data = subset(lead_counties_five, contb_receipt_amt > 0)) +
  facet_wrap( ~ county) +
    scale_fill_manual(values=party_color) +
  geom_bar(stat = "identity", position = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Five Candidates with Highest Total Donation Dollars")
p9b
#Rather intersting how much support that Carson has in Multnomah county. 
#Considering how "Blue" the area is thought of I am suprised to see so much 
#support for Carson.

```

```{r final eval graph 3}
# 
top_five_only <- subset(cont_table.fc_by_summarize, cand_nm == "Sanders, Bernard" |
                    cand_nm == "Carson, Benjamin S." |
                    cand_nm == "Clinton, Hillary Rodham" |
                    cand_nm == "Cruz, Rafael Edward 'Ted'" |
                    cand_nm == "Paul, Rand")

mth_sort2 = factor(top_five_only$mth_yr, levels = c("Apr-15","May-15","Jun-15","Jul-15","Aug-15","Sep-15"))

p9c <- ggplot(aes(x = mth_sort2, y = monthly_total, fill = cand_party),
              data = subset(top_five_only)) +
  facet_wrap( ~ cand_nm) +
  scale_fill_manual(values = party_color) +
  geom_bar(stat = "identity", position = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p9c
#We can see that the support varies some from month to month, especially
#a large upswing in July for Clinton, it would be worth investigating what
#caused such a massive upswing that month.

```
## Reflection

Oregon has long been a Blue State when it comes to the Presidential elections,
so finding that the Democratic Candidates had a greater support base was not
unexpected.  I was also unsurprised to see that the top counties and zip codes
are those that make up and surround the Portland area as it is the most 
population dense.  It was interesting to see that even in these urban areas
there was still a strong showing for Republican candidates, which I found 
to be unexpected.  Overall the information presented matched expected results
and was an interesting peak into the financial side of the election.